---
title: "Progress Report"
author: "Demetri Pananos"
date: "18/11/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

```{r}
# source('2018-09 Apixiban Bayesian Models/Run Models/Model_7.R')
```



# Introduction

Our goal is two fold:

* We want to accurately predict concentrations of apixaban over time 

* We want to understand the effect of covariates on the concentration

To do this, I have opted for a Bayesian approach.  The model posits that the concentration as a function of time looks like

$$ C(t) = \dfrac{D}{V}\dfrac{k_a}{k_a - k}\Big(\exp(-k_at) - \exp(-kt) \Big) \>.$$

Here, $D$ is the dose size, $V$ is the volume of the absorptive compartment, and $k_a$ and $k$ are rates of absorption and exrection, resptively.  Aside from the dose size, these parameters are unknown and will need to be estimated.  I opt for a heirarchal model which allows each patient to have their own set of paramters and thus their own curve. The model posits that each patient's parameters are draws from a population distirbution whos mean depends on the covariates.  Model details are discussed below.


# Data

Data for model construction is provided by Dr. Tirrona.  The data consist of 36 patients sampled over 8 time points. Two covariates are measured: Sex, and Disease status (NAFLD and Control).  Shown below are the mean concentrations over the four possible combinations of covariate levels.

```{r}
f.plot
```

Female patients have higher concentration, on average, than male patients.  There is little difference between control and NAFLD groups in female patients.  There appears to be an effect of NAFLD on concentration for male patients.  We should be cautious about this effect, as the male control group consits of 2 patients.



